Use and acquisition of causal relations for decision making under uncertainty using imperfect information repeated games

### 2018-06-23 Propuesta/Proposal_English.tex
metodologia experimental

A particular case of a \textit{higher level structure}; i.e., beyond associative patterns, is the case of \textit{causal} structure. A causal structure encondes a series of \textit{cause-effect} relations between events.

@@

@@

	Methods based in Deep Neural Networks aren't supposed to explain why a certain output was produced since those methods are based in \textit{parallel distributed representations} and the same goes for any other learning algorithm that uses Deep Neural Networks, like Deep Reinforcement learning; when AlphaGo (\cite{silver2017mastering}) defeated the world-champion, humanity didn't learn anything new about the game of Go, because the algorithm was not designed to explain its moves, it was just curve fitting.

	Learning a causal model of an environment and using it to act upon the environment allows to \textit{explain} aspects of the model that a purely associative model would not be abel to explain. It allows to ask \textit{why}.

@@

	\subsection{Research Question}
	How a rational decision maker who faces an uncertain environment which is governed by a causal mechanism can learn and make use of this causal structure in order to make good choices? How can causal structure help a decision maker in order to guide his learning process?
	\subsection{Hypothesis}
	Let $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. Then, if a decision maker doesn't know the probabilities of the uncertain events in $E$, by repeatedly making decisions he can learn causal information and use it to find the optimal actions (in the sense of expected utility theory) in less, or equal, rounds than if he doesn't consider causal information.
	\subsection{Objectives}
		\subsubsection{General Objectives}
			The proposed research has as a general objective to find how a rational autonomous agent who faces an uncertain environment can discover and use causal information in order to make good choices that point towards a concise objective such as the maximization of some function.
		\subsubsection{Specific Objectives}
		In order to achieve the General Objective, we need to separate it into smaller problems which are required to be solved first.

@@

Notice that none of the previous works consider any kind of causal structure. For this kind of problems, causal relations between actions and consequences could be considered, in this way an action could be conceived as an \textit{intervention} over some environment, which is in fact how humans consider actions in the world (\cite{hagmayer2009decision}). Consider a set of variables that are causally related between them and to a \textit{reward} variable whose value has to be maximized by intervening one of the variables. It is shown by \cite{lattimoreNIPS2016} that adding causal information in a fixed budget decision problem allows the decision maker to learn \textit{faster} had he not considered causal information. Their work requires that the causal model is fully known to the decision maker, this requirement is relaxed later by \cite{sen2017identifying} who requires only that some part of the causal model is known and allow interventions over the unknown part.

@@

@@

\subsection{Belief updating}

@@

The three cases that we notice are:
\begin{itemize}
\item The decision maker knows the causal model.
\item The decision maker knows only the graphical \textit{structure} of the causal model.
\item The decision maker knows nothing about the causal model. 
\end{itemize}
In the following we show how the first two cases, which are particular cases of our general problem, can be solved and we show that the proposed solution for the first two cases sheds light on how the general case could be solved.
\subsubsection{Fully knowing the causal graphical model}
If the decision maker knows the causal graphical model, then he could easily calculate the effect on the target variable that any of his actions has and then choose the action that has the highest probability of causing a \textit{desired} value of the target variable. The decision thus made is the one that maximizes the expected utility, by construction. Also, choosing this action is the \textit{best response} action, and therefore a Nash equilibrium for the game. 
\subsubsection{Knowing only the graph structure}

\subsubsection{General case: Model unknown}

\subsection{Validation of the proposed methods}

\subsection{Experimental methodology}
\subsubsection{Internal Validation: Reproducibility}
\subsubsection{External Validation: Different problems}

\subsection{Working plan}
Concrete steps towards solving the problem:
\begin{itemize}
\item Literature review, understanding the problem, considering several models and possible solutions.
\item Formalize the model. 
\end{itemize}
\subsubsection{Publication plan}
\begin{itemize}
\item Gonzalez-Soto, M., Sucar, L.E., Escalante, H.J., \textbf{Playing against Nature: causal discovery for decision making under uncertainty}. Accepted for a poster presentation at the CausalML 2018 workshop at ICML / IJCAI / AAMAS 2018.
\end{itemize}

@@

In this way, the decision maker will have beliefs about the CPT's in the form of parameters of several Dirichlet distributions. Using the agent's current beliefs, a causal graphical model can be specified. Using this fully specified (structure + parameters) as a true model, the decision maker will make her choice as in Case 1. When the decision maker observes the value of the target variable, she will update the parameters that specify her beliefs.

>>>>

A particular case of a \textit{higher level structure}; i.e., beyond associative patterns, is the case of \textit{causal} structure. A causal structure encondes a series of \textit{cause-effect} relations between events and knowing such relations allows a decision maker to add extra knowledge into the uncertainty of his environment and also allows to plan ahead his actions since he can predict what a certain action will cause (\cite{spirtes2000causation}, \cite{pearl2018theoretical}).

@@

	Also, an extra condition, known as Causal Sufficiency, about the nature of the model itself is considered: for any variable $X$ in the model $\mathcal{G}$ there are no causes of $X$ outside of the model $\mathcal{G}$ (\cite{spirtes2000causation}, \cite{pearl2009causality}, \cite{sucar2015probabilistic}).

@@

	Methods based in Deep Neural Networks aren't supposed to explain why a certain output was produced since those methods are based in \textit{parallel distributed representations} and the same goes for any other learning algorithm that uses Deep Neural Networks, like Deep Reinforcement learning; when AlphaGo (\cite{silver2017mastering}) defeated the world-champion, humanity didn't learn anything new about the game of Go, because the algorithm was not designed to explain its moves, it was just curve fitting, although a very sofisticated method for achieving it. Using only associations between variables it is not possible to infer which one is the cause and which one the effect, something extra is required.

	On the other hand, learning a causal model of an environment and using it to act upon the environment allows to \textit{explain} aspects of the model that a purely associative model would not be abel to explain. It allows to ask \textit{why}.

@@

	\subsection{Research Question}
	How a rational decision maker who faces an uncertain environment which is governed by a causal mechanism can learn and make use of this causal structure in order to make good choices? How can causal structure help a decision maker in order to guide his learning process?
	\subsection{Hypothesis}
	Let $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. Then, if a decision maker doesn't know the probabilities of the uncertain events in $E$, by repeatedly making decisions he can learn causal information and use it to find the optimal actions (in the sense of expected utility theory) in less, or equal, rounds than if he doesn't consider causal information.
	\subsection{Objectives}
		\subsubsection{General Objectives}
			The proposed research has as a general objective to find how a rational autonomous agent who faces an uncertain environment can discover and use causal information in order to make good choices that point towards a concise objective such as the maximization of some function.
		\subsubsection{Specific Objectives}
		In order to achieve the General Objective, we need to separate it into smaller problems which are required to be solved first. In first place, it is required to find a general way of defining a game that captures the interaction between a decision maker and a causal-governed environment.

		In second place, to find a family of distributions that represent any past knowledge and current beliefs about the causal structure of the environment and a way to use these beliefs in order to make choices according to the decision maker's preferences.

		Beliefs must be updated after external information is provided by the environment in response to any given action that was taken by the decision maker. A general updating criteria must be provided which uses the causal nature of the environment.

@@

Notice that none of the previous works consider any kind of causal structure. For this kind of problems, causal relations between actions and consequences could be considered, in this way an action could be conceived as an \textit{intervention} over some environment, which is in fact how humans consider actions in the world (\cite{hagmayer2009decision}) and that human beings use and modify causal knowledge during a sequential decision making process (\cite{hagmayer2013repeated}). Consider a set of variables that are causally related between them and to a \textit{reward} variable whose value has to be maximized by intervening one of the variables. It is shown by \cite{lattimoreNIPS2016} that adding causal information in a fixed budget decision problem allows the decision maker to learn \textit{faster} had he not considered causal information. Their work requires that the causal model is fully known to the decision maker, this requirement is relaxed later by \cite{sen2017identifying} who requires only that some part of the causal model is known and allow interventions over the unknown part.

@@

We are considering each game as a learning round, where the only information that is carried from a game to another is what is learned by the decision maker via a process of belief updating. The causal model is assumed to remain fixed and each initialization at the beginning of each new game is independent from any past outcome.

@@

\subsection{Belief updating}
The whole idea is that a decision maker is able to learn from his environment by observing, and reasoning about, the effects of his actions. The observed information will modify his current knowledge, which is encoded as probabilistic beliefs, about the environment in a way that is similar to how human beings modify causal knowledge while intervening in the world (\cite{hagmayer2013repeated}).

If a decision maker encodes his current knowledge, and ignorance, about some relevant characteristic of his environment, then it is known that belief updating using Bayes' Theorem is the only way of updating that is coherent with the rational preference axioms (\cite{bernardo2000bayesian}).

Bayesian updating is also of great importance in the theory of Learning in Games (\cite{fudenberg1998theory}) since it is known (\cite{shoham2008multiagent}) that if all the players in a game perfom Bayesian updating, then the plays generated by these measures converge to the plays that would be generated if all players shared their knowledge.

Even though we know that Bayesian updating is the correct \textit{framework} for belief updating, it remains unanswered how to concretely use causal information in order to update the parameters that control the distributions used to express beliefs in a tractable way.

@@

The three cases that we notice are:
\begin{itemize}
\item The decision maker knows the causal model.
\item The decision maker knows only the graphical \textit{structure} of the causal model.
\item The decision maker knows nothing about the causal model. 
\end{itemize}
In the following we show how the first two cases, which are particular cases of our general problem, can be solved and we show that the proposed solution for the first two cases sheds light on how the general case could be solved.
\subsubsection{Fully knowing the causal graphical model}
If the decision maker knows the causal graphical model, then he could easily calculate the effect on the target variable that any of his actions has and then choose the action that has the highest probability of causing a \textit{desired} value of the target variable. The decision thus made is the one that maximizes the expected utility, by construction. Also, choosing this action is the \textit{best response} action, and therefore a Nash equilibrium for the game. 
\subsubsection{Knowing only the graph structure}
If a decision maker knows the structure of the graph, but he does not know the parameters, then he encodes this ignorance, along with any useful previous knowledge, as a probability distribution over a suitable space. In the preliminary results that will be shown later a particular case is considered, where the beliefs have the form of a Dirichlet distribution for each of the Conditional Probability Tables that characterize a Bayesian Network.

In general, what must be specified is a way of encoding current causal knowledge into a probability distribution in such a way that a \textit{local} (i.e. in each round) causal model can be obtained from current beliefs in order to make a decision using current causal knowledge. 
\subsubsection{General case: Model unknown}

\subsection{Validation of the proposed methods}
In order to show that an agent has succeeded in acquiring a causal model from his environment via interacting with it and using it in order to make good choices, 
\subsection{Experimental methodology}
The proposed methodology will be tested in the following way: A fully specified causal model, which satisfies the conditiones required above, will be given and used as a ground truth. The causal model will not be available to the decision maker, since one of the basic assuptions is that the agent does not know the causal model that controls his environment. The true causal model will be used to simulate realizations of the other variables in the distribution given any action taken by the decision maker.

One, or more, of the variables is to be selected as \textit{available} to interventions by the decision maker, the set of possible values for the selected variable determines the set of available actions of the decision maker in the formulation as a game. Also, another variable is selected as a \textit{target} variable, preferably an endogenous variable in the model in order to be affected by any action taken by the agent. The set of possible values of this variable must be in correspondance with the set of outcomes of the decision problem that the agent faces so the value that the target variable outputs has an effect over the preferences of the agent.

A distribution with initial parameters $\theta_0$ will be specified.

Given a number $T$ of rounds, and inside round $t \in \{1,2,...,T \}$ the agent will make an action $a \in \mathcal{A}$ and then observe the state of the other variables of the model. Using this information, current beliefs will be updated according to the criteria to be established.

The chosen actions will be compared to classic on-line decision making 
\subsubsection{Internal Validation: Reproducibility}
For the results obtained by following the methodology described in previous sections be reproducible, the causal model used to simulate an environment must be fully specified by describing the probabilistic parameters that control it. The initial distributions given to the decision maker to encode his beliefs must be shown as well with the updating criteria used when information from the environment becomes available. 
\subsubsection{External Validation: Different problems}
For the proposed method to be tested in different decision problems, it is required that the correspondent causal model satisfies all of the conditions previously described, and any derived algorithm will be designed in such a way that the characteristics of the causal model are an external input. In this way, if an external user is able to give a causal model, the method is expected to work independent of the particularities of the causal model which is being tested.
\subsection{Working plan}
Concrete steps towards solving the problem:
\begin{itemize}
\item Literature review, understanding the problem, considering several models and possible solutions.
\item Formalize the model. 
\end{itemize}
\subsubsection{Publication plan}
\begin{itemize}
\item Gonzalez-Soto, M., Sucar, L.E., Escalante, H.J., \textbf{Playing against Nature: causal discovery for decision making under uncertainty}. Accepted for a poster presentation at the CausalML 2018 workshop at ICML / IJCAI / AAMAS 2018. July 15 2018, Stockholm, Sweden.
\end{itemize}

@@

The decision maker will have beliefs about the CPT's in the form of parameters of several Dirichlet distributions. Using the agent's current beliefs, a causal graphical model can be specified. Using this fully specified (structure + parameters) as a true model, the decision maker will make his choice as in Case 1. When the decision maker observes the value of the target variable, he will update the parameters that specify his beliefs.
